Supp mat: time series paper

### 2017-06-29 writing/manuscript_draft.Rmd
added sensitivity analysis to alpha and beta

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987]. Simple population models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker). In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series (cite). Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized it for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organims, that 15 years was adequate to assess longterm trends in abundance. However, these studies, and other past work, have focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression.
<!-- In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.  -->

@@

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here statistical power is simply the fraction of simulations with significant trends in abundance from simple linear regression. Then, the minimum time required is the minimum time required to obtain statistical power above some pre-determined threshold (historically at 0.8).

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). A power of 0.8 would imply that given a trend exists, there is a 0.8 probability of detecting the trend. Complementary, there would be a 0.2 probability of a false negative.

<!-- example of how subsampling and empirical approach works -->
```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Harp Seals over time. The line is the best fit line from linear regresssion. (b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
# Example of subsampling empirical data
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n
source('scripts/calculate_power_metric.R')
    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)

>>>>

For time series data, two general approaches can be used. First, as can be done with experimental data, we can relax the definition of statistical power (cite Bolker). Instead of using classic statistical tests (e.g. t-test), and corresponding formulas for determinig sample size, simulation approaches can be used [@Gerrodette1987]. Simple models can be simulated with parameter values corresponding to a population of interest. Out of a large number of simulations, power can be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. Then, time series simulations of different lengths are run to determine the minimum time series length required to be confident that you will be able to answer your question (cite Bolker). In addition to using simulated data, empirical time series can also be used. We usually do not have multiple replicates of similar populations, but we can subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized amodel specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organim, 15 years was adequate to assess longterm trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression. We examine 878 population time series to estimate this minimum time series length required.
<!-- In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.  -->

@@

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here statistical power is simply the fraction of simulations with significant trends in abundance. A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression. Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). <!--In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years. Therefore, 16 years would be our estimate for $T_{min}$.-->

### 2017-06-29 writing/manuscript_draft.tex
added sensitivity analysis to alpha and beta

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). Rhodes and Jonzen (2011) examined the optimal allocation of
effort between spatial and temporal replicates. Using simple populations
models, they found that the allocation of effort depends on
environmental variation, spatial and temporal autocorrelation, and
observer error. Rueda-Cediel et al. (2015) also used a modeling
approach, but parameterized it for a threatened snail,
\emph{Tasmaphena lamproides}. They found that for this short-lived
organims, that 15 years was adequate to assess longterm trends in
abundance. However, these studies, and other past work (citation), have
focused either on theoretical aspects of monitoring design or focused on
single species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. Here, we determine the minimum time series length to
determine if a population is incresing or decreasing according to a
significant slope coefficient in simple linear regression.

>>>>

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Giron-Nava et
al. 2017). For example, Rhodes and Jonzen (2011) examined the optimal
allocation of effort between spatial and temporal replicates. Using
simple populations models, they found that the allocation of effort
depends on environmental variation, spatial and temporal
autocorrelation, and observer error. Rueda-Cediel et al. (2015) also
used a modeling approach, but parameterized it for a threatened snail,
\emph{Tasmaphena lamproides}. They found that for this short-lived
organims, that 15 years was adequate to assess longterm trends in
abundance. However, these studies, and other past work, have focused
either on theoretical aspects of monitoring design or focused on single
species.

In this paper, we use both simulations and a database of empirical time
series to determine the minimum number of years required to address
various questions. More specifically, we estimate the minimum time
series length required to asssess longterm changes in abundance via
simple linear regression.

### 2017-06-29 writing/supp_mat_draft.Rmd
added sensitivity analysis to alpha and beta

This supplementary material includes:

1. Detailed example of subsampling and power calculations
2. Additional figures
3. Figures and results from other questions of interest
4. Additional references

@@

## Detailed example of subsampling and power calculations

@@

We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample. We can call each set of subsamples, of the same length, a set. For each subsample, we conduct linear regression and extract model coefficients and p-values. Then, the fraction of subsamples within a set that show significant trends (signfiiant slope coefficient) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.

@@

Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statisical power of 0.8 implies...

@@

```{r, echo=F}
# show an example of subsampling and power calculation routines
```

## Additional results from the main manuscript

```{r, echo=F}
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```

We then correlated minimum time required for each population with its corresponding life history characteristics. In figure \ref{fig:biological_correlates} we examined miminmum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days). None of these variables had much explanatory power in accounting for the variance in the minimum time required. see table \ref{table:model_output}}

@@

@@

## Results from other questions of interest

### Simulations with more complicated population model

In the main text, we showed how a simple population model could be simulated repeatidly to estimate the power obtained with time series of increasing length. The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required. This model is purely phenemilogical and does not include any species life history. Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model, the (MODEL NAME HERE):

@@

We simulate this model repetively for different lengths of time series. Statistical power is the fraction of time series of a set length that showed signicant trends in abundance, given a true trend in abundance. *How do I determine significance here? Through linear regression again, or do I look at non-zero estimation of the parameter values?*

### Empirical approach with more complicated population model (use model similar to de Valpine work?)

### Growth rate calculations

Instead of detecting a trend over time with linear regression, we could also calculate the geometric growth rate of the population. In figure \ref{fig:growth_rate}, we show how to calculate growth rates for subsamples of a time series. First, we created subsamples of each possible length from the full 35 year time series, as we did in the main next. Next, we calculated the mean and standard deviation of growth rates for each possible time series length (Fig. \ref{fig:growth_rate}b). Lastly, we calculated the percent error $\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|$ between the mean of each time series length (observed) and the overall population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c), we show the percent error as a function of time series length. Here we define the minimum time required as the minimum number of years to achieve less than 20% error.

@@

### Using Generalized additive model to identify significant trends

In the main text, we examined the minimum time required to identify a trend in abundance via linear regression. This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit. Generalized additive models (GAMs) are more general than general linear models and allow more flexible. GAMs are models where a response variable depends on unknown smooth functions of explanatory variables. GAMs, therefore, can identify relationships between response variables and explantory variables that are non-linear and perhaps more complicated. The downside of GAMs is they typically require more data and are also prone to overfitting (citations).

@@

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
# show gam example side-by-side plot
plot(1,1)
```

@@

### Minimum time to detect density-dependence in time series

For decades, the importance of density-dependence in natural populations has been an important question in ecology (citations). To address this question, @Knape2012 used data from the Global Population Dynamics Database [@GPDD2010]. They fitted 627 time series with a Gompertz population model and found evidence for density-dependence in 45% of the time series. Each time series in their database contained at least 15 datapoints, and ranged from 15 to 157. A follow up question to their work would be, what time series length is required to confidently estimate density-dependence.

\clearpage
### References

>>>>

<!--
This supplementary material includes:

1. Detailed example of subsampling and power calculations
  - figure of subsampling routine
2. Additional results
  - statistical model results of main text
  - biological correlates plot
  - minimum time required and biological class
  - sensitivity of alpha and power
3. Figures and results from other questions of interest
  - simulations with more realistic population model
  - empirical example with more realistic population model
  - minimum time required to estimate longterm geometric growth rate
  - empirical approach with longterm geometric growth rate
  - estimating minimum time required for generalized additive models

4. Additional references
-->

\tableofcontents

@@

# Detailed example of subsampling and power calculations

@@

We first extract all contiguous subsamples of the time series. This leads to 34 two-year subsamples, 33 two-year subsamples, and so forth until a single 35-year subsample. We can call each set of subsamples, of the same length, a set. For each subsample, we conduct linear regression and extract model coefficients and p-values. Then, the fraction of subsamples within a set that show significant trends (significant slope coefficient) is the statistical power. It is important to note that we only consider subsamples to be significant if they are significant in the same direction as the complete 35-year time series.

@@

Then, we determine an appropriate level of statistical power that we find acceptable. Traditionally, this has been at 0.8, however, this is purely historical. Statistical power of 0.8 implies...

@@

<!-- example of how subsampling and empirical approach works -->
```{r, echo=F,message=F,warning=F,fig.cap='(a) Population size of Harp Seals over time. The line is the best fit line from linear regresssion. (b) Statistical power for different subsets of the time series in panel a.\\label{fig:empirical_approach_example}'}
# Example of subsampling empirical data
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='13192') #13192 for Harp seals
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n
source('scripts/calculate_power_metric.R')
    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)

@@

In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years. Therefore, 16 years would be our estimate for $T_{min}$.

# Additional results from the main manuscript

## Statistical model of minimum time required and time series characteristics

@@

## Minimum time required and biological correlates
```{r, echo=F}
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```

We then correlated minimum time required for each population with its corresponding life history characteristics. In figure \ref{fig:biological_correlates} we examined minimum time required versus generation length (years), litter size (n), adult body mass (grams), maximum longevity (years), egg mass (grams), and incubation (days). None of these variables had much explanatory power in accounting for the variance in the minimum time required. see table \ref{table:model_output}}

## Minimum time series and biological class

When we examined differences in the minimum time required there were similar patterns with biological classes. For instance, more time is required for species within the Actinopterygii class compared to other species (Fig. \ref{fig:class}a). These differences between biological classes can be explained by differences in population variability, with species in Actinopterygii have larger variability in population size from year-to-year.

```{r,echo=F,fig.cap='(a) Minimum time required to estimate change in abundance for species class, (b) longterm trend (estimated slope coefficient) by species class, (c) interannual variability in population size by species class, and (d) temporal autocorrelation by species class.\\label{fig:class}',fig.height=6}

# Pull in data
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
# Biological correlates plots
par(mfrow=c(2,2),oma=c(8,1,1.5,0.5),mar=c(0,4,0,1))

boxplot(pop_info$min_time_for_power~pop_info$class, lwd = 1, ylab = 'minimum time required',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
stripchart(pop_info$min_time_for_power~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))

boxplot(abs(pop_info$overall_trend)~pop_info$class, lwd = 1, ylab = ' ',col='white',cex.lab=1.4,cex.axis=1.2,las=2,xaxt="n")
stripchart(abs(pop_info$overall_trend)~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
  mtext('| trend strength |',2,line=3.5,cex=1.2)

boxplot(pop_info$variance~pop_info$class, lwd = 1, ylab = 'population variance',col='white',cex.lab=1.4,cex.axis=1.2,las=2)
stripchart(pop_info$variance~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))

boxplot(pop_info$autocorrelation~pop_info$class, lwd = 1, ylab = 'autocorrelation',col='white',cex.lab=1.4,cex.axis=1.2,las=2)
stripchart(pop_info$autocorrelation~pop_info$class, vertical = TRUE,method = "jitter", add = TRUE, pch = 20, col = rgb(0.5,0.5,0.5,0.5))
```

## Sensitivity analysis of $\alpha$ and power

Estimates of $T_{min}$ depend strongly on the values used for the significance level ($\alpha$) and the probability of type II error ($\beta$), both of which are set by the practitioner. Here we explore how estimates of $T_{min}$ are affected by changes in each of these parameters. We see that the minimum time required increases with increases in statistical power or decreases with increases in the threshold for statistical significance (Fig \ref{fig:min_time_vs_alpha_beta}).

```{r,echo=F,message=F,error=F,fig.cap='Minimum time required to assess longterm trends in abundance for values of statistical significance ($\\alpha$) and power ($1-\\beta$).\\label{fig:min_time_vs_alpha_beta}'}
# Bring in theoretical code
# Build plot of $T_min$ vs alpha and another versus beta
# Explain results after building figures
setwd("~/Desktop/Research/time-series-project")
source('scripts/calculate_slope.R') 
source('scripts/theoretical-models/simulate_pop.R')
# calculate_power = function(r,phi,sigma,trials,max.time,alpha){
#   trials   = trials
#   max.time = max.time
#   r        = r
#   phi      = phi
#   sigma    = sigma
#   
#   y=replicate(n=trials,simulate_pop(r,phi,sigma,max.time)) 
#   
#   power=sum((apply(y,2,calculate_p_value)<alpha) & sign(apply(y,2,calculate_slope))==sign(r))/trials 
#   
#   return(power)
# }
# 
# 
# min_time_needed= function(alpha,power){
#   years_to_sample=seq(2,30,by=1)
#   y=lapply(X=years_to_sample,FUN = calculate_power,phi =0.5,r=0.5,sigma = 1,trials=100,alpha=alpha)
#   min_time_needed = years_to_sample[tail(which(y<power),1)] + 1
#   return(min_time_needed)
# }
# 
# alpha_values=seq(0.001,0.2,by=0.01)
# power_values=seq(0.65,0.95,by=0.01)
# parameter_combinations = expand.grid(alpha_values,power_values)
# names(parameter_combinations) = c('alpha','power')
# parameter_combinations$min_time_required = NA
# 
# for (i in 1:nrow(parameter_combinations)){
# 
#   parameter_combinations$min_time_required[i] =  min_time_needed(parameter_combinations[i,1],parameter_combinations[i,2])
#  
#   #print(i)
# }

#write.csv(parameter_combinations,file = 'analysis-outputs/min_time_vs_alpha_and_power.csv',quote = F,row.names = F)

# Pull in data to prevent R from running for a long time to create data from scratch
parameter_combinations = read.csv('analysis-outputs/min_time_vs_alpha_and_power.csv',header=T)

require(ggplot2)
ggplot(parameter_combinations, aes(x = alpha, y = power, z = min_time_required, fill=min_time_required)) +
  geom_tile() + 
  labs(title = " ", x = expression(alpha), y = expression(paste('power ',(1-beta))), fill = expression(T[min])) +
  scale_fill_distiller(palette="Spectral", na.value="white") +
  theme_bw()
```

# Results from other questions of interest

## Simulations with more complicated population model

In the main text, we showed how a simple population model could be simulated repeatedly to estimate the power obtained with time series of increasing length. The model in the main text simulated linear population growth with only a slope coefficient, y-intercept, and noise parameter required. This model is purely phenomenological and does not include any species life history. Here, we use the same routine as the main text, but simulate from a more biologically-realistic population model, the (MODEL NAME HERE):

@@

```{r,echo=F}
#Include simulations of more interesting model here
```

We simulate this model repetitively for different lengths of time series. Statistical power is the fraction of time series of a set length that showed significant trends in abundance, given a true trend in abundance. *How do I determine significance here? Through linear regression again, or do I look at non-zero estimation of the parameter values?*

## Empirical approach with more complicated population model (use model similar to de Valpine work?)

## Growth rate calculations

Instead of detecting a trend over time with linear regression, we could also calculate the geometric growth rate of the population. In figure \ref{fig:growth_rate}, we show how to calculate growth rates for subsamples of a time series. First, we created subsamples of each possible length from the full 35 year time series, as we did in the main next. Next, we calculated the mean and standard deviation of growth rates for each possible time series length (Fig. \ref{fig:growth_rate}b). Lastly, we calculated the percent error $\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|$ between the mean of each time series length (observed) and the overall population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c), we show the percent error as a function of time series length. Here we define the minimum time required as the minimum number of years to achieve less than 10% error.

@@

## Using Generalized additive model to identify significant trends

In the main text, we examined the minimum time required to identify a trend in abundance via linear regression. This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit. Generalized additive models (GAMs) are more general than general linear models and allow more flexible. GAMs are models where a response variable depends on unknown smooth functions of explanatory variables. GAMs, therefore, can identify relationships between response variables and explanatory variables that are non-linear and perhaps more complicated. The downside of GAMs is they typically require more data and are also prone to overfitting (citations).

@@

```{r,echo=F,message=F,warning=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
# show gam example side-by-side plot
 setwd("~/Desktop/Research/time-series-project")
 load("cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")
 source('scripts/calculate_power_metric_gam.R')
  par(mfrow=c(1,2),oma=c(4,4,0.5,0.5))
    pop = subset(long_dat,long_dat$ID=='7558') #13192 for Harp seals, 7558 for Thunnus obesus
    plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='time (years)',pch=16,las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
    mtext(text='population size', side=2,line=4,cex=1.2)
# could insert all possible regressions of length n

    power_example=calculate_power_for_subsamples(pop$popvalue,0.05)
    plot(power_example,cex.axis=1.2,cex.lab=1.2,ylab='statistical power',xlab='years sampled',las=1,pch=16)
    mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
# could also insert standard error bars
    abline(h=0.8,col='red',lwd=2,lty=2)
    abline(v=head(which(power_example>0.8),1),col='red',lwd=2,lty=2)
```

@@

## Minimum time to detect density-dependence in time series

For decades, the importance of density-dependence in natural populations has been an important question in ecology (citations). To address this question, @Knape2012 used data from the Global Population Dynamics Database [@GPDD2010]. They fitted 627 time series with a Gompertz population model and found evidence for density-dependence in 45% of the time series. Each time series in their database contained at least 15 data points, and ranged from 15 to 157. A follow up question to their work would be, what time series length is required to confidently estimate density-dependence.

# References
